---
title: "Stock Return Analysis: Permutation Tests & Portfolio Value at Risk"
subtitle: "Apple Inc. (AAPL) vs Microsoft Corp. (MSFT) — 2021–2024"
author: "Statistical Finance Report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center"
)
set.seed(42)
```

---

# 1. Introduction

This report analyses daily stock prices of **Apple Inc. (AAPL)** and **Microsoft Corp. (MSFT)** over the period **2021–2024** (approximately 4 years of trading data). The study covers:

1. Data download and computation of logarithmic daily returns.
2. Permutation-based hypothesis tests on covariance, variance equality, and equality of expected returns.
3. Portfolio profit analysis: buying 100 shares of each stock and selling 100 trading days later.
4. Comparison of the empirical profit distribution with a fitted normal distribution.
5. Empirical and parametric (normal) **Value at Risk (VaR)** at the 5%, 1%, and 0.1% levels.

---

# 2. Data & Returns

## 2.1 Packages and Data Download

```{r packages-data}
# Install missing packages automatically
required_packages <- c("quantmod", "ggplot2", "gridExtra", "moments", "knitr", "kableExtra", "qqplotr", "scales")
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE))
    install.packages(pkg, repos = "https://cloud.r-project.org")
}

library(quantmod)
library(ggplot2)
library(gridExtra)
library(moments)
library(knitr)
library(kableExtra)
library(qqplotr)
library(scales)

# Download adjusted closing prices from Yahoo Finance
start_date <- "2021-01-01"
end_date   <- "2024-12-31"

getSymbols("AAPL", src = "yahoo", from = start_date, to = end_date, auto.assign = TRUE)
getSymbols("MSFT", src = "yahoo", from = start_date, to = end_date, auto.assign = TRUE)

price_aapl <- as.numeric(Ad(AAPL))
price_msft <- as.numeric(Ad(MSFT))
dates      <- index(Ad(AAPL))

cat(sprintf("AAPL: %d observations  |  MSFT: %d observations\n",
            length(price_aapl), length(price_msft)))
cat(sprintf("Date range: %s  to  %s\n", min(dates), max(dates)))
```

> **Note:** We use **adjusted closing prices** (accounting for dividends and splits) to ensure comparability across time.

## 2.2 Price Chart

```{r price-chart, fig.height=4}
# Index prices to 100 at the start date so relative performance is directly comparable.
# Raw prices differ in level (AAPL ~$130 vs MSFT ~$220 at start), which would make
# one series appear visually flatter even if its percentage moves are identical.
df_prices <- data.frame(
  date         = rep(dates, 2),
  price        = c(price_aapl, price_msft),
  price_indexed = c(price_aapl / price_aapl[1] * 100,
                    price_msft / price_msft[1] * 100),
  ticker       = rep(c("AAPL", "MSFT"), each = length(dates))
)

ggplot(df_prices, aes(x = date, y = price_indexed, colour = ticker)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100, linetype = "dotted", colour = "grey50") +
  scale_colour_manual(values = c("AAPL" = "#1f77b4", "MSFT" = "#d62728")) +
  labs(title    = "Relative Performance: AAPL vs MSFT (2021–2024)",
       subtitle = "Both series indexed to 100 at 2021-01-04 — shows percentage growth, not price level",
       x = "Date", y = "Indexed Price (start = 100)", colour = "Ticker") +
  theme_minimal(base_size = 12)
```

## 2.3 Logarithmic Daily Returns

Daily log-returns are defined as:

$$r_t = \ln\!\left(\frac{P_t}{P_{t-1}}\right)$$

Log-returns are preferred over simple returns because they are **time-additive** and tend to be closer to normally distributed for short horizons.

```{r returns}
ret_aapl  <- diff(log(price_aapl))
ret_msft  <- diff(log(price_msft))
ret_dates <- dates[-1]
n         <- length(ret_aapl)

# Summary table
summary_df <- data.frame(
  Ticker = c("AAPL", "MSFT"),
  N      = c(n, n),
  Mean   = c(mean(ret_aapl), mean(ret_msft)),
  SD     = c(sd(ret_aapl),   sd(ret_msft)),
  Min    = c(min(ret_aapl),  min(ret_msft)),
  Max    = c(max(ret_aapl),  max(ret_msft)),
  Skewness = c(skewness(ret_aapl), skewness(ret_msft)),
  ExKurtosis = c(kurtosis(ret_aapl) - 3, kurtosis(ret_msft) - 3)
)

kable(summary_df, digits = 6,
      caption = "Descriptive Statistics of Daily Log-Returns",
      col.names = c("Ticker","N","Mean","Std Dev","Min","Max","Skewness","Ex. Kurtosis")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r returns-chart, fig.height=4}
df_ret <- data.frame(
  date   = rep(ret_dates, 2),
  ret    = c(ret_aapl, ret_msft) * 100,
  ticker = rep(c("AAPL","MSFT"), each = n)
)

ggplot(df_ret, aes(x = date, y = ret, colour = ticker)) +
  geom_line(alpha = 0.55, linewidth = 0.4) +
  scale_colour_manual(values = c("AAPL" = "#1f77b4", "MSFT" = "#d62728")) +
  labs(title  = "Daily Log-Returns (%) — AAPL and MSFT",
       x = "Date", y = "Return (%)", colour = "Ticker") +
  theme_minimal(base_size = 12)
```

---

# 3. Permutation Tests

Permutation (randomisation) tests are **non-parametric** inference procedures. They require no distributional assumptions. The key idea is:

> Under the null hypothesis the observed test statistic should be no more extreme than what we would expect by **random reassignment** of the data.

We use **N = 10,000 permutations** for all tests. All p-values are two-sided.

```{r perm-setup}
N_PERM   <- 10000
combined <- c(ret_aapl, ret_msft)   # pooled sample (used in tests 3b and 3c)
```

---

## 3.1 Test of Covariance Significance

### Hypotheses

$$H_0: \text{Cov}(r_{\text{AAPL}},\, r_{\text{MSFT}}) = 0 \quad \text{(returns are linearly independent)}$$

$$H_1: \text{Cov}(r_{\text{AAPL}},\, r_{\text{MSFT}}) \neq 0 \quad \text{(returns are linearly associated)}$$

**Test statistic:** $T = \widehat{\text{Cov}}(r_{\text{AAPL}},\, r_{\text{MSFT}})$

Under $H_0$ the pairing between the two series is arbitrary, so we break it by randomly permuting one series. The permutation distribution approximates the sampling distribution of the covariance under $H_0$.

```{r perm-cov, fig.height=4}
obs_cov  <- cov(ret_aapl, ret_msft)
perm_cov <- replicate(N_PERM, cov(sample(ret_aapl), ret_msft))
p_val_cov <- mean(abs(perm_cov) >= abs(obs_cov))

# Shade bars that are at least as extreme as the observed statistic (two-sided).
# This makes the p-value visually intuitive: shaded area / total area = p-value.
df_perm_cov <- data.frame(cov = perm_cov)
df_perm_cov$extreme <- abs(df_perm_cov$cov) >= abs(obs_cov)

ggplot(df_perm_cov, aes(x = cov, fill = extreme)) +
  geom_histogram(bins = 80, colour = "white", alpha = 0.9) +
  scale_fill_manual(values = c("FALSE" = "#aec6cf", "TRUE" = "#e63946"),
                    labels = c("FALSE" = "Not as extreme", "TRUE" = "As extreme or more"),
                    name   = "Relative to observed") +
  geom_vline(xintercept =  obs_cov, colour = "darkred", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = -obs_cov, colour = "darkred", linewidth = 1.3, linetype = "dashed") +
  annotate("text", x = obs_cov, y = Inf,
           label = sprintf("  obs = %.2e", obs_cov),
           vjust = 2, hjust = -0.05, colour = "darkred", size = 3.8, fontface = "bold") +
  labs(title    = sprintf("Permutation Distribution of Covariance  (p = %.4f)", p_val_cov),
       subtitle = "Red bars = permuted values at least as extreme as observed (two-sided); dashed lines = ± observed",
       x = "Covariance (permuted)", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r cov-result}
cat(sprintf("Observed covariance:  %.2e\n", obs_cov))
cat(sprintf("Permutation p-value:  %.4f\n", p_val_cov))
cat(sprintf("Decision (alpha=0.05): %s\n",
            ifelse(p_val_cov < 0.05, "REJECT H0", "FAIL TO REJECT H0")))
```

### Interpretation

The observed covariance is **`r formatC(obs_cov, format="e", digits=2)`**. Under H₀ (returns are independent), permuting the time-pairing of one series should produce covariances scattered symmetrically around zero — which is exactly what the histogram shows. The proportion of permuted covariances at least as extreme as the observed one (red bars) is the p-value: **p = `r round(p_val_cov, 4)`**.

Since p `r ifelse(p_val_cov < 0.05, "< 0.05, we **reject H₀**", "> 0.05, we **fail to reject H₀**")`. `r if(p_val_cov < 0.05) "In all 10,000 random reshufflings of the data, not a single permuted covariance was as large as the observed value — the observed covariance lies completely outside the permutation distribution. This is overwhelming evidence that the co-movement of AAPL and MSFT returns is not a chance artefact. Economically, both are large-cap US tech stocks exposed to the same macro drivers (interest rates, risk sentiment, sector rotation), so a positive dependence is expected." else "The observed covariance falls within the bulk of the permutation distribution, so we cannot distinguish it from what random chance would produce under independence."`

---

## 3.2 Test of Equality of Variances

### Hypotheses

$$H_0: \sigma^2_{\text{AAPL}} = \sigma^2_{\text{MSFT}} \quad \text{(equal return volatility)}$$

$$H_1: \sigma^2_{\text{AAPL}} \neq \sigma^2_{\text{MSFT}} \quad \text{(different volatilities)}$$

**Test statistic:** $T = \ln\!\left(\hat\sigma^2_{\text{AAPL}} / \hat\sigma^2_{\text{MSFT}}\right)$

We use the **log-ratio** of sample variances (rather than the raw ratio) because it is symmetric under $H_0$ and has a more stable permutation distribution. Under $H_0$, pooling both samples and randomly splitting them into two equal groups should produce a ratio close to 1 (i.e., log-ratio close to 0).

```{r perm-var, fig.height=4}
obs_var_ratio  <- var(ret_aapl) / var(ret_msft)
perm_var_ratio <- replicate(N_PERM, {
  perm <- sample(combined)
  var(perm[1:n]) / var(perm[(n+1):(2*n)])
})
p_val_var <- mean(abs(log(perm_var_ratio)) >= abs(log(obs_var_ratio)))

df_perm_var <- data.frame(lr = log(perm_var_ratio))
df_perm_var$extreme <- abs(df_perm_var$lr) >= abs(log(obs_var_ratio))

ggplot(df_perm_var, aes(x = lr, fill = extreme)) +
  geom_histogram(bins = 80, colour = "white", alpha = 0.9) +
  scale_fill_manual(values = c("FALSE" = "#c8e6c9", "TRUE" = "#e63946"),
                    labels = c("FALSE" = "Not as extreme", "TRUE" = "As extreme or more"),
                    name   = "Relative to observed") +
  geom_vline(xintercept =  log(obs_var_ratio), colour = "darkred", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = -log(obs_var_ratio), colour = "darkred", linewidth = 1.3, linetype = "dashed") +
  annotate("text", x = log(obs_var_ratio), y = Inf,
           label = sprintf("  obs = %.3f", log(obs_var_ratio)),
           vjust = 2, hjust = -0.05, colour = "darkred", size = 3.8, fontface = "bold") +
  labs(title    = sprintf("Permutation Distribution of log(Var Ratio)  (p = %.4f)", p_val_var),
       subtitle = "Red bars = permuted values at least as extreme as observed (two-sided); dashed lines = ± observed",
       x = "log(var_AAPL / var_MSFT) — permuted", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r var-result}
cat(sprintf("Observed variance — AAPL: %.2e\n", var(ret_aapl)))
cat(sprintf("Observed variance — MSFT: %.2e\n", var(ret_msft)))
cat(sprintf("Variance ratio (AAPL/MSFT): %.4f\n", obs_var_ratio))
cat(sprintf("Permutation p-value:        %.4f\n", p_val_var))
cat(sprintf("Decision (alpha=0.05): %s\n",
            ifelse(p_val_var < 0.05, "REJECT H0", "FAIL TO REJECT H0")))
```

### Interpretation

The observed variance ratio is **`r round(obs_var_ratio, 4)`** (log-ratio: `r round(log(obs_var_ratio), 4)`), meaning AAPL daily returns are `r ifelse(obs_var_ratio > 1, "more", "less")` volatile than MSFT by a factor of `r round(max(obs_var_ratio, 1/obs_var_ratio), 4)`. Under H₀ the two series have equal variance, so randomly pooling all observations and splitting them back into two groups of equal size should produce a log-ratio close to zero. The red bars in the histogram show how often random splits yielded a log-ratio at least as large in absolute value as the one we observed.

The permutation p-value is **p = `r round(p_val_var, 4)`**. `r ifelse(p_val_var < 0.05, "Since p < 0.05, we **reject H₀** — the difference in volatility between the two stocks is statistically significant and is unlikely to have arisen by chance. In practical terms, AAPL carries more daily return risk than MSFT over this period.", "Since p > 0.05, we **fail to reject H₀** — the data do not provide sufficient evidence to distinguish the two volatilities. We cannot rule out that the observed ratio is due to random variation.")`

---

## 3.3 Test of Equality of Expected Returns

### Hypotheses

$$H_0: \mu_{\text{AAPL}} = \mu_{\text{MSFT}} \quad \text{(equal mean daily returns)}$$

$$H_1: \mu_{\text{AAPL}} \neq \mu_{\text{MSFT}} \quad \text{(different mean daily returns)}$$

**Test statistic:** $T = \bar r_{\text{AAPL}} - \bar r_{\text{MSFT}}$

Under $H_0$ the label "AAPL" vs "MSFT" is exchangeable; randomly reassigning observations to groups should produce differences near zero.

```{r perm-mean, fig.height=4}
obs_diff_mean  <- mean(ret_aapl) - mean(ret_msft)
perm_diff_mean <- replicate(N_PERM, {
  perm <- sample(combined)
  mean(perm[1:n]) - mean(perm[(n+1):(2*n)])
})
p_val_mean <- mean(abs(perm_diff_mean) >= abs(obs_diff_mean))

df_perm_mean <- data.frame(dm = perm_diff_mean)
df_perm_mean$extreme <- abs(df_perm_mean$dm) >= abs(obs_diff_mean)

ggplot(df_perm_mean, aes(x = dm, fill = extreme)) +
  geom_histogram(bins = 80, colour = "white", alpha = 0.9) +
  scale_fill_manual(values = c("FALSE" = "#ffe0b2", "TRUE" = "#e63946"),
                    labels = c("FALSE" = "Not as extreme", "TRUE" = "As extreme or more"),
                    name   = "Relative to observed") +
  geom_vline(xintercept =  obs_diff_mean, colour = "darkred", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = -obs_diff_mean, colour = "darkred", linewidth = 1.3, linetype = "dashed") +
  annotate("text", x = obs_diff_mean, y = Inf,
           label = sprintf("  obs = %.2e", obs_diff_mean),
           vjust = 2, hjust = -0.05, colour = "darkred", size = 3.8, fontface = "bold") +
  labs(title    = sprintf("Permutation Distribution of Mean Difference  (p = %.4f)", p_val_mean),
       subtitle = "Red bars = permuted values at least as extreme as observed (two-sided); dashed lines = ± observed",
       x = "Mean(AAPL) − Mean(MSFT) — permuted", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r mean-result}
cat(sprintf("Mean return AAPL: %.6f  (%.4f%% daily)\n",
            mean(ret_aapl), mean(ret_aapl)*100))
cat(sprintf("Mean return MSFT: %.6f  (%.4f%% daily)\n",
            mean(ret_msft), mean(ret_msft)*100))
cat(sprintf("Observed difference:      %.2e\n", obs_diff_mean))
cat(sprintf("Permutation p-value:      %.4f\n", p_val_mean))
cat(sprintf("Decision (alpha=0.05): %s\n",
            ifelse(p_val_mean < 0.05, "REJECT H0", "FAIL TO REJECT H0")))

# Classical Welch t-test for comparison
tt <- t.test(ret_aapl, ret_msft)
cat(sprintf("Welch t-test p-value (classical):  %.4f\n", tt$p.value))
```

### Interpretation

The difference in sample means is **`r formatC(obs_diff_mean, format="e", digits=2)`** per day (`r round(obs_diff_mean * 252 * 100, 2)`% annualised). Under H₀, pooling both return series and randomly splitting them should produce differences scattered around zero — which the bell-shaped permutation distribution confirms. The permutation p-value is **p = `r round(p_val_mean, 4)`** (Welch t-test: p = `r if(!is.null(tt)) round(tt$p.value, 4)`; the two approaches agree closely).

`r ifelse(p_val_mean < 0.05, "Since p < 0.05, we **reject H₀** — the mean daily returns differ significantly. However, note that statistical significance here does not imply economic significance: the difference is tiny in absolute terms and would be overwhelmed by transaction costs in practice.", "Since p > 0.05, we **fail to reject H₀** — the observed difference in mean returns is consistent with random variation. This is the typical result for large-cap stocks over horizons of a few years: daily mean returns are estimated with very high variance, making the test low-powered.")`

---

## 3.4 Summary of Permutation Tests

```{r perm-summary-table}
perm_summary <- data.frame(
  Test         = c("Covariance significance",
                   "Equality of variances",
                   "Equality of expected returns"),
  H0           = c("Cov = 0", "σ²_AAPL = σ²_MSFT", "μ_AAPL = μ_MSFT"),
  Statistic    = c(round(obs_cov, 6),
                   round(obs_var_ratio, 4),
                   round(obs_diff_mean, 6)),
  p_value      = c(p_val_cov, p_val_var, p_val_mean),
  Decision     = c(ifelse(p_val_cov  < 0.05, "Reject H₀", "Fail to reject"),
                   ifelse(p_val_var  < 0.05, "Reject H₀", "Fail to reject"),
                   ifelse(p_val_mean < 0.05, "Reject H₀", "Fail to reject"))
)

kable(perm_summary, digits = 4,
      caption = "Permutation Test Results (N = 10,000, α = 0.05)",
      col.names = c("Test","Null Hypothesis H₀","Observed Statistic","p-value","Decision (α=0.05)")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  column_spec(5, bold = TRUE,
              color = ifelse(perm_summary$p_value < 0.05, "red", "darkgreen"))
```

---

# 4. Portfolio Profit Analysis

## 4.1 Strategy Description

For every trading day $t$ we **buy 100 shares of each stock** at the closing price and **sell them 100 trading days later**:

$$\Pi_t = 100 \cdot (P^{\text{AAPL}}_{t+100} - P^{\text{AAPL}}_t) + 100 \cdot (P^{\text{MSFT}}_{t+100} - P^{\text{MSFT}}_t)$$

Transaction costs are ignored. The result is a time series of overlapping 100-day holding-period profits.

```{r portfolio}
N_SHARES  <- 100
HORIZON   <- 100

buy_idx   <- 1:(length(price_aapl) - HORIZON)
sell_idx  <- buy_idx + HORIZON

profit_aapl      <- N_SHARES * (price_aapl[sell_idx] - price_aapl[buy_idx])
profit_msft      <- N_SHARES * (price_msft[sell_idx] - price_msft[buy_idx])
profit_portfolio <- profit_aapl + profit_msft
n_port           <- length(profit_portfolio)

port_summary <- data.frame(
  Metric = c("N (windows)", "Mean profit (USD)", "Median profit (USD)",
             "Std Dev (USD)", "Min (USD)", "Max (USD)",
             "Skewness", "Excess Kurtosis"),
  Value  = c(n_port,
             round(mean(profit_portfolio), 2),
             round(median(profit_portfolio), 2),
             round(sd(profit_portfolio), 2),
             round(min(profit_portfolio), 2),
             round(max(profit_portfolio), 2),
             round(skewness(profit_portfolio), 4),
             round(kurtosis(profit_portfolio) - 3, 4))
)

kable(port_summary, caption = "Descriptive Statistics — Portfolio Profit") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

---

## 4.2 Profit Distribution vs Normal

```{r profit-histogram, fig.height=7}
mu_port  <- mean(profit_portfolio)
sd_port  <- sd(profit_portfolio)
x_seq    <- seq(min(profit_portfolio) - 500, max(profit_portfolio) + 500, length.out = 600)
df_norm  <- data.frame(x = x_seq, y = dnorm(x_seq, mu_port, sd_port))

bw <- (max(profit_portfolio) - min(profit_portfolio)) / 60

alpha_levels <- c(0.05, 0.01, 0.001)
var_emp  <- quantile(profit_portfolio, probs = alpha_levels)
var_norm <- qnorm(alpha_levels, mean = mu_port, sd = sd_port)

# --- Panel 1: Empirical VaR ---
# Each VaR level gets its own distinct colour from a colourblind-friendly palette.
emp_lines <- data.frame(
  x     = var_emp,
  level = factor(c("5%", "1%", "0.1%"), levels = c("5%", "1%", "0.1%"))
)

p1 <- ggplot(data.frame(profit = profit_portfolio), aes(x = profit)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = bw,
                 fill = "#ffcc99", colour = "white", alpha = 0.9) +
  geom_line(data = df_norm, aes(x = x, y = y),
            colour = "darkblue", linewidth = 1.2, linetype = "solid") +
  geom_vline(data = emp_lines, aes(xintercept = x, colour = level),
             linewidth = 1.2) +
  geom_label(data = emp_lines,
             aes(x = x, y = Inf, label = paste0("VaR\n", level), colour = level),
             vjust = 1.3, size = 3.2, fill = "white", label.size = 0, show.legend = FALSE) +
  scale_colour_manual(values = c("5%" = "#2166ac", "1%" = "#d6604d", "0.1%" = "#762a83"),
                      name = "Empirical VaR") +
  labs(title    = "Panel A — Empirical VaR",
       subtitle = sprintf("Solid blue = fitted N(μ=%.0f, σ=%.0f)", mu_port, sd_port),
       x = "Profit (USD)", y = "Density") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

# --- Panel 2: Empirical vs Normal VaR comparison ---
# Pair each level's empirical and normal VaR side by side using different linetypes.
both_lines <- data.frame(
  x      = c(var_emp, var_norm),
  level  = rep(factor(c("5%", "1%", "0.1%"), levels = c("5%", "1%", "0.1%")), 2),
  method = rep(c("Empirical", "Normal"), each = 3)
)

p2 <- ggplot(data.frame(profit = profit_portfolio), aes(x = profit)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = bw,
                 fill = "#ffcc99", colour = "white", alpha = 0.9) +
  geom_line(data = df_norm, aes(x = x, y = y),
            colour = "darkblue", linewidth = 1.2) +
  geom_vline(data = both_lines,
             aes(xintercept = x, colour = level, linetype = method),
             linewidth = 1.1) +
  scale_colour_manual(values = c("5%" = "#2166ac", "1%" = "#d6604d", "0.1%" = "#762a83"),
                      name = "VaR level") +
  scale_linetype_manual(values = c("Empirical" = "solid", "Normal" = "dashed"),
                        name = "Method") +
  labs(title    = "Panel B — Empirical vs Normal VaR",
       subtitle = "Solid = empirical quantile, Dashed = normal quantile; colour = level",
       x = "Profit (USD)", y = "Density") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

grid.arrange(p1, p2, nrow = 2)
```

### Shapiro–Wilk Normality Test

We formally test whether portfolio profits follow a normal distribution.

**Hypotheses:**

$$H_0: \text{Portfolio profit is normally distributed}$$
$$H_1: \text{Portfolio profit is NOT normally distributed}$$

```{r shapiro}
sw <- shapiro.test(sample(profit_portfolio, min(5000, n_port)))
cat(sprintf("Shapiro–Wilk test: W = %.5f,  p-value = %.4e\n", sw$statistic, sw$p.value))
cat(sprintf("Decision (alpha=0.05): %s\n",
            ifelse(sw$p.value < 0.05, "REJECT H0 — non-normal", "FAIL TO REJECT H0 — consistent with normal")))
```

### Q–Q Plot

```{r qqplot, fig.height=4.5}
# The grey band is a 95% pointwise confidence envelope for the normal Q-Q plot.
# Points inside the band are consistent with normality; points outside indicate departures.
ggplot(data.frame(profit = profit_portfolio), aes(sample = profit)) +
  stat_qq_band(distribution = "norm",
               dparams = list(mean = mu_port, sd = sd_port),
               fill = "#b0c4de", alpha = 0.4) +
  stat_qq_line(distribution = "norm",
               dparams = list(mean = mu_port, sd = sd_port),
               colour = "red", linewidth = 1.1) +
  stat_qq_point(distribution = "norm",
                dparams = list(mean = mu_port, sd = sd_port),
                alpha = 0.3, size = 0.9, colour = "#333333") +
  labs(title    = "Q–Q Plot: Portfolio Profit vs Normal Distribution",
       subtitle = "Grey band = 95% pointwise confidence envelope; points outside → departure from normality",
       x = "Theoretical Normal Quantiles (USD)", y = "Empirical Quantiles (USD)") +
  theme_minimal(base_size = 12)
```

### Interpretation

`r if(sw$p.value < 0.05) paste0("The Shapiro–Wilk test strongly rejects normality (p = ", formatC(sw$p.value, format='e', digits=2), "). The Q–Q plot confirms this: the tails deviate notably from the reference line, indicating **heavier tails** than a Gaussian distribution. This is consistent with the excess kurtosis of ", round(kurtosis(profit_portfolio)-3, 3), " and skewness of ", round(skewness(profit_portfolio), 3), ". These features — common in financial return data — mean that extreme losses (and gains) occur more frequently than a normal model predicts.") else paste0("The Shapiro–Wilk test does not reject normality at the 5% level (p = ", round(sw$p.value, 4), "). The Q–Q plot is broadly consistent with a normal distribution.")`

---

# 5. Value at Risk (VaR)

**Value at Risk at level $\alpha$** is the threshold loss such that the probability of the portfolio return falling *below* it is $\alpha$:

$$\text{VaR}_\alpha = Q_\alpha(\Pi) \quad \text{i.e.,} \quad P(\Pi \leq \text{VaR}_\alpha) = \alpha$$

A *negative* VaR denotes a loss. We compute two variants:

- **Empirical VaR:** the empirical $\alpha$-quantile of observed profits.  
- **Normal VaR:** the $\alpha$-quantile of $\mathcal{N}(\hat\mu, \hat\sigma^2)$ fitted to the data.

```{r var-table}
var_df <- data.frame(
  Level         = c("5%", "1%", "0.1%"),
  Empirical_VaR = round(var_emp, 2),
  Normal_VaR    = round(var_norm, 2),
  Difference    = round(var_emp - var_norm, 2)
)

kable(var_df, caption = "Value at Risk — Empirical vs Normal (USD)",
      col.names = c("Confidence Level", "Empirical VaR (USD)",
                    "Normal VaR (USD)", "Difference (Emp − Norm)")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r var-viz, fig.height=4}
var_comp <- data.frame(
  level  = rep(c("5%","1%","0.1%"), 2),
  method = rep(c("Empirical","Normal"), each = 3),
  value  = c(var_emp, var_norm)
)
var_comp$level <- factor(var_comp$level, levels = c("5%","1%","0.1%"))

ggplot(var_comp, aes(x = level, y = value, fill = method)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = sprintf("$%.0f", value),
                vjust = ifelse(value < 0, 1.4, -0.4)),
            position = position_dodge(width = 0.6), size = 3.2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("Empirical" = "#e63946", "Normal" = "#457b9d")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  annotate("text", x = 0.5, y = min(c(var_emp, var_norm)) * 0.85,
           label = "← Loss territory", hjust = 0, size = 3.2, colour = "grey40") +
  labs(title    = "Empirical vs Normal VaR by Confidence Level",
       subtitle = "Negative values = losses; bars below zero are worse outcomes",
       x = "Significance Level (α)", y = "VaR (USD)", fill = "Method") +
  theme_minimal(base_size = 12)
```

### Interpretation

| Level | Empirical VaR | Normal VaR | Observation |
|-------|-------------|-----------|-------------|
| **5%**   | `r round(var_emp[1],0)` USD | `r round(var_norm[1],0)` USD | `r ifelse(abs(var_emp[1]) > abs(var_norm[1]), "Emp. loss larger — normal underestimates risk", "Normal loss larger — normal overestimates risk")` |
| **1%**   | `r round(var_emp[2],0)` USD | `r round(var_norm[2],0)` USD | `r ifelse(abs(var_emp[2]) > abs(var_norm[2]), "Emp. loss larger — normal underestimates risk", "Normal loss larger")` |
| **0.1%** | `r round(var_emp[3],0)` USD | `r round(var_norm[3],0)` USD | `r ifelse(abs(var_emp[3]) > abs(var_norm[3]), "Emp. loss larger — normal significantly underestimates tail risk", "Normal loss larger")` |

`r if(abs(var_emp[3]) > abs(var_norm[3])) "The divergence grows as α decreases (i.e., as we move deeper into the left tail). This is the **fat-tail effect**: extreme losses occur more frequently in real data than a Gaussian model predicts. Using the normal VaR at the 0.1% level would therefore **underestimate** the true tail risk, which is a well-known pitfall in financial risk management." else "The normal approximation performs reasonably well across all confidence levels for this portfolio."`

---

# 6. Conclusions

```{r conclusions-table}
conclusions <- data.frame(
  Question = c(
    "Are AAPL and MSFT returns linearly dependent?",
    "Do they have equal volatility?",
    "Do they have equal expected returns?",
    "Is the portfolio profit normally distributed?",
    "Does normal VaR capture tail risk accurately?"
  ),
  Finding = c(
    ifelse(p_val_cov  < 0.05,
           paste0("Yes — covariance is significant (p = ", round(p_val_cov, 4), ")"),
           paste0("No evidence — covariance not significant (p = ", round(p_val_cov, 4), ")")),
    ifelse(p_val_var  < 0.05,
           paste0("No — variances differ significantly (p = ", round(p_val_var, 4), ")"),
           paste0("Possibly — no significant difference (p = ", round(p_val_var, 4), ")")),
    ifelse(p_val_mean < 0.05,
           paste0("No — means differ significantly (p = ", round(p_val_mean, 4), ")"),
           paste0("Possibly — no significant difference (p = ", round(p_val_mean, 4), ")")),
    ifelse(sw$p.value < 0.05,
           paste0("No — Shapiro–Wilk rejects normality (p = ", formatC(sw$p.value, format='e', digits=1), ")"),
           paste0("Consistent with normal (p = ", round(sw$p.value, 4), ")")),
    ifelse(abs(var_emp[3]) > abs(var_norm[3]),
           "No — normal VaR underestimates extreme losses (fat tails)",
           "Approximately yes — normal VaR is close to empirical")
  )
)

kable(conclusions, caption = "Summary of Key Findings",
      col.names = c("Question", "Finding")) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  column_spec(1, bold = TRUE, width = "40%") |>
  column_spec(2, width = "60%")
```

> **Key takeaway for risk management:** Assuming normality of portfolio returns leads to systematic underestimation of tail risk at very low probability levels. Empirical or parametric heavy-tailed distributions (e.g., Student-$t$, Cornish–Fisher expansion) should be preferred for VaR estimation at the 1% and 0.1% levels.

---

*Report generated with R `r R.version$major`.`r R.version$minor` · quantmod · ggplot2 · moments*
